---
title: "Download and format templates for OOM"
author: "David M. Kaplan"
format: html
editor: visual
---

## Overview

This script contains R code to download the base IGN Lidar, OSM and ESRI imagery files to use as templates for creating the orienteering map in OpenOrienteeringMapper. Use instructions below.

### Steps to use this script

1) First install [R](https://cran.r-project.org/) and [RStudio](https://posit.co/products/open-source/rstudio/)

2) Next edit the `parameters` and `ign-filenames` chunks of the script to suit your area. Link to IGN Lidar terrain data can be identified [here](https://cartes.gouv.fr/telechargement/IGNF_MNT-LIDAR-HD)

3) Open this file and RStudio and click `Render`

4) The script will slowly install all necessary packages, download baseline data and create files to use as templates for OpenOrienteeringMapper

## Setup

### Parameters


```{r}
#| label: parameters
# Directories for storing downloads
odir = "templates"
cdir = file.path(odir,"contours")
hdir = file.path(odir,"hillshades")

# For dividing Lidar height data
height_breaks = c(-Inf,-1,0,1,2,5,100,Inf)

# List of color names for height categories
height_cols0 = c("subterranean"= "orange",
                 "ground"="tan",
                 "low"="yellow",
                 "med"="lightgreen",
                 "high"="darkgreen",
                 "lofty"="white",
                 "NA"="gray")

# Convert to numerical RGB values
height_cols = lapply(height_cols0,col2rgb)

# bounding box of actual area we want to keep
# These limits are so big, they cut nothing so map will have extent of IGN Lidar tiles
bb = st_bbox(c(xmin=-10,xmax=10,
               ymin=30,ymax=50),
             crs=4326)
```

```{r}
#| label: ign-filenames
# From website: https://cartes.gouv.fr/telechargement/IGNF_MNT-LIDAR-HD
mnt_fns = c(
  "https://data.geopf.fr/wms-r?SERVICE=WMS&VERSION=1.3.0&EXCEPTIONS=text/xml&REQUEST=GetMap&LAYERS=IGNF_LIDAR-HD_MNT_ELEVATION.ELEVATIONGRIDCOVERAGE.LAMB93&FORMAT=image/geotiff&STYLES=&CRS=EPSG:2154&BBOX=756999.75,6276000.25,757999.75,6277000.25&WIDTH=2000&HEIGHT=2000&FILENAME=LHD_FXX_0757_6277_MNT_O_0M50_LAMB93_IGN69.tif",
  "https://data.geopf.fr/wms-r?SERVICE=WMS&VERSION=1.3.0&EXCEPTIONS=text/xml&REQUEST=GetMap&LAYERS=IGNF_LIDAR-HD_MNT_ELEVATION.ELEVATIONGRIDCOVERAGE.LAMB93&FORMAT=image/geotiff&STYLES=&CRS=EPSG:2154&BBOX=756999.75,6275000.25,757999.75,6276000.25&WIDTH=2000&HEIGHT=2000&FILENAME=LHD_FXX_0757_6276_MNT_O_0M50_LAMB93_IGN69.tif",
  "https://data.geopf.fr/wms-r?SERVICE=WMS&VERSION=1.3.0&EXCEPTIONS=text/xml&REQUEST=GetMap&LAYERS=IGNF_LIDAR-HD_MNT_ELEVATION.ELEVATIONGRIDCOVERAGE.LAMB93&FORMAT=image/geotiff&STYLES=&CRS=EPSG:2154&BBOX=755999.75,6275000.25,756999.75,6276000.25&WIDTH=2000&HEIGHT=2000&FILENAME=LHD_FXX_0756_6276_MNT_O_0M50_LAMB93_IGN69.tif",
  "https://data.geopf.fr/wms-r?SERVICE=WMS&VERSION=1.3.0&EXCEPTIONS=text/xml&REQUEST=GetMap&LAYERS=IGNF_LIDAR-HD_MNT_ELEVATION.ELEVATIONGRIDCOVERAGE.LAMB93&FORMAT=image/geotiff&STYLES=&CRS=EPSG:2154&BBOX=755999.75,6276000.25,756999.75,6277000.25&WIDTH=2000&HEIGHT=2000&FILENAME=LHD_FXX_0756_6277_MNT_O_0M50_LAMB93_IGN69.tif"
)

mnh_fns = gsub("MNT","MNH",mnt_fns)
```

### Install & load necessary packages

The code below tries to load necessary packages. If they are absent, it tries to install them.

```{r}
#| label: setup
# List of necessary packages
pkgs.cran = c("tidyverse","sf","stars","basemaps","cols4all","foreach",
              "OpenStreetMap","osmdata","hillshader")
pkgs.github = c()

# Check if CRAN packages installed
# If not, install them
for (pkg in c(pkgs.cran)) {
  message("Attempting to load package: ",pkg)
  
  if (!require(pkg,character.only=TRUE)) {
    message("Missing package ",pkg,"! Attempting install...")
    install.packages(pkg,repos="https://cloud.r-project.org")
  }
  
  library(pkg,character.only=TRUE)
}

# Github packages
for (pkgg in pkgs.github) {
  pkg = basename(pkgg)
  message("Attempting to load package: ",pkg)
  
  if (!require(pkg,character.only=TRUE)) {
    message("Missing package ",pkg,"! Attempting install from github...")
    remotes::install_github(pkgg,upgrade=FALSE)
  }

  library(pkg,character.only=TRUE)
}
```

```{r}
#| label: helper-functions
# Gets CRS of UTM system that best corresponds to data
lonlat2UTM = function(lonlat) {
  utm = (floor((lonlat[1] + 180) / 6) %% 60) + 1
  if (lonlat[2] > 0) {
    utm + 32600
  } else{
    utm + 32700
  }
}

linear_rescale = function(m,new_min=0,new_max=255) {
  mn = min(m)
  mx = max(m)
  
  new_min + (m - mn) / (mx-mn) * (new_max - new_min)
}

quantile_cut = function(m,n,probs=seq(0,1,length.out=n+1)) {
  q = quantile(m,probs=probs)
  m = cut(m,breaks=unique(q),include.lowest=TRUE)
  return(list(m=m,q=q))
}
```

```{r}
#| label: create-directories
dir.create(odir,recursive = TRUE,showWarnings = FALSE)
dir.create(cdir,recursive = TRUE,showWarnings = FALSE)
dir.create(hdir,recursive = TRUE,showWarnings = FALSE)
```

## Download IGN Lidar products

!!!DONE TO HERE!!!! !!!!NEED TO ADD CROP HERE!!!!

```{r}
#| label: download-terrain-files
#| eval: !expr "!file.exists(file.path(odir,'mnt.RDS'))"

fns = tempfile(as.character(1:length(mnt_fns)),fileext=".tif")

for (n in seq_along(mnt_fns)) {
  download.file(mnt_fns[n],destfile=fns[n])  
}

x = st_mosaic(fns)
y = read_stars(x)

saveRDS(y,file.path(odir,"mnt.RDS"))
```

```{r}
#| label: download-height-files
#| eval: !expr "!file.exists(file.path(odir,'mnh.RDS'))"

fns = tempfile(as.character(1:length(mnt_fns)),fileext=".tif")

for (n in seq_along(mnh_fns)) {
  download.file(mnh_fns[n],destfile=fns[n])  
}

x = st_mosaic(fns)
y = read_stars(x)

saveRDS(y,file.path(odir,"mnh.RDS"))
```

## Download tiles

```{r}
#| label: utm-epsg-bbox
y = readRDS(file.path(odir,"mnt.RDS"))

# Bounding box with some added edge for tiles download
b = st_bbox(y) + c(-200,-200,200,200) # Add some around edge
bll = b |> st_transform(4326) # For downloading high res image

# Find best UTM coordinate system for area
bc = bll |> st_as_sfc() |> st_centroid()
utm = bc |> st_coordinates() |> lonlat2UTM()
```


```{r}
#| label: download-tiles
#| eval: !expr "!file.exists(file.path(odir,'esri.high-res.tif'))"

# OpenStreetMap
o = basemap_stars(b,"osm","streets",1)
o |> write_stars(file.path(odir,"osm.tif"),
            options="COMPRESS=LZW",type="Byte")
o |> st_warp(crs = utm) |>
  write_stars(file.path(odir,"osm.utm.tif"),
            options="COMPRESS=LZW",type="Byte")

# ESRI satellite imagery
es = basemap_stars(b,"esri","world_imagery",1)
es |> write_stars(file.path(odir,"esri.tif"),
            options="COMPRESS=LZW",type="Byte")
es |> st_warp(crs=utm) |>
  write_stars(file.path(odir,"esri.utm.tif"),
            options="COMPRESS=LZW",type="Byte")

# Higher res version of ESRI satellite imagery using OpenStreetMap package
om = openmap(bll[4:3],bll[2:1],type="esri-imagery",zoom=18) |>
  st_as_stars()
om |> write_stars(file.path(odir,"esri.high-res.tif"),
                  options="COMPRESS=LZW",type="Byte")
om |> st_warp(crs=utm) |> 
  write_stars(file.path(odir,"esri.high-res.utm.tif"),
              options="COMPRESS=LZW",type="Byte")
```

```{r}
#| label: osmdata
#| eval: !expr "!file.exists(file.path(odir,'osm.osm'))"
#| 
# Get vector data from OSM
x = opq(bll) 
osmdata_xml(x, file.path(odir,"osm.osm"))
```


## Contours and hillshading

```{r}
#| label: contours-terrain
y = readRDS(file.path(odir,"mnt.RDS"))

# Downsample data to get somewhat smoother contour lines
y4 = st_downsample(y,3,FUN=mean) 

get_alt_breaks = function(x,step) {
  r = range(x[[1]])
  seq(floor(r[1]/step)*step,ceiling(r[2]/step)*step,by=step)
}

# Fine contours for potential visualization
co2.5 = st_contour(y,breaks=get_alt_breaks(y,2.5),contour_lines = TRUE)

# Smoother 25 m contours
st25 = get_alt_breaks(y4,25)
co25 = st_contour(y4,breaks=st25,contour_lines = TRUE)

# Smoother 5 m contours without 25 m contours
st5 = get_alt_breaks(y4,5)
st5.no.25 = setdiff(st5,st25)

co5.no.25 = st_contour(y4,breaks=st5.no.25,contour_lines = TRUE)

# Write to shapefiles
st_write(co2.5,file.path(odir,"contours","2.5m.shp"),delete_dsn=TRUE)
st_write(co25,file.path(odir,"contours","25.0m.smooth.4.shp"),delete_dsn=TRUE)
st_write(co5.no.25,file.path(odir,"contours","5.0m.no.25m.smooth.4.shp"),delete_dsn=TRUE)
```


```{r}
#| label: hillshading
#| eval: !expr "!file.exists(file.path(odir,'mnt.hillshade.tif'))"
 
y = readRDS(file.path(odir,"mnt.RDS"))

hs = hillshader(as(y,"Raster"),shader="ray_shade") |>
  st_as_stars()
hs[[1]] = round(hs[[1]]*255)
write_stars(hs,file.path(odir,"mnt.hillshade.tif"),
            options="COMPRESS=LZW",type="Byte")
```

## Analyze height data

```{r}
#| label: classify-heights
#| eval: !expr "!file.exists(file.path(odir,'mnh.classified.tif'))"

y = readRDS(file.path(odir,"mnh.RDS"))

# First, cut into classes
y2 = y
y2[[1]] = cut(y2[[1]],breaks=height_breaks,
              labels=names(height_cols))

# Convert values to colors
z2 = st_apply(y2,1:2,\(.) height_cols[[.]])
# This has RGB band in first dimension spot, but this seems to
# not affect saving to geotiff. This can be fixed by splitting first
# dimension to attributes (split function) and then merging attributes
# (merge function).

# Write at geotiff
write_stars(z2,file.path(odir,"mnh.classified.tif"),
            options="COMPRESS=LZW",type="Byte")
```

```{r}
#| label: quantile-classify-heights
#| eval: !expr "!file.exists(file.path(odir,'mnh.quantile-classified.tif'))"

y = readRDS(file.path(odir,"mnh.RDS"))

n = 10
p = c4a("brewer.paired",n=n)
phex = col2rgb(p)

yy = y
qq = yy[[1]] |> quantile_cut(n=n)
yy[[1]] = qq$m
yy = yy |> st_apply(1:2,\(.) phex[,.])

# Write at geotiff
write_stars(yy,file.path(odir,"mnh.quantile-classified.tif"),
            options="COMPRESS=LZW",type="Byte")
```



```{r}
#| label: downsample-heights
#| eval: !expr "!file.exists(file.path(odir,'mnh.downsample_hillshade.tif'))"

y = readRDS(file.path(odir,"mnh.RDS"))

# Down sample to 3 m scale
ymn = st_downsample(y,5,FUN=mean) 
ysd = st_downsample(y,5,FUN=sd)

hs = hillshader(as(ymn,"Raster"),shader="ray_shade") |>
  st_as_stars()
hs[[1]] = round(hs[[1]]*255)
write_stars(hs,file.path(odir,"mnh.downsample_hillshade.tif"),
            options="COMPRESS=LZW",type="Byte")

hs = hillshader(as(ysd,"Raster"),shader="ray_shade") |>
  st_as_stars()
hs[[1]] = round(hs[[1]]*255)
write_stars(hs,file.path(odir,"mnh.downsample_sd_hillshade.tif"),
            options="COMPRESS=LZW",type="Byte")

```

## Download North Arrow image

```{r}
#| label: download-north-arrow
download.file("https://storage.needpix.com/thumbs/magnetic-33208_1280.png",
              file.path(odir,"magnetic-33208_1280.png"))

n = read_stars(file.path(odir,"magnetic-33208_1280.png"))
n = n |> st_rgb()

# Replace black by blue
n[[1]][n[[1]][]=="#000000FF"] = "#00FFFFFF"

n2 = st_apply(n,1:2,\(.) as.integer(col2rgb(.,alpha=TRUE)))

write_stars(n2,file.path(odir,"magnetic-33208_1280.blue.tif"),
            options="COMPRESS=LZW",type="Byte")
```

